The prime of their life - the halflife of acting talent

### 2017-09-02 01_halflife-of-acting-talent/article.Rmd
formatting changes

@@

  html = glue::glue("

  .finding {
    background-color:#FF6A5C;
    height: 100px;
    color: #FFF;
    display: flex;
    padding-top: 10px;
    padding-bottom: 10px;
    border-radius: 25px;
    align-items: center;
    text-align:center;
  }

  .finding .sidebar { 
    width: 20%;
    font-weight: bold;
    font-size: 14pt;
  }

  .finding .body {
    width: 80%
  }

    <div class='finding'>

      <div class='sidebar'><img src='https://github.com/TMBish/lab_posts/blob/master/00_utilities/lightbulb.png?raw=true'><span> Finding!</span></div>

      <div class='body'> <span> <<finding_text>> </span> </div>

    </div>

    <br>" %>% str_replace_all("\\n", ""), .open = "<<", .close = ">>")

@@

dict %>% pandoc.table()

@@

films %>% 
  filter(title %in% c("The Departed", "Titanic", "No Country for Old Men", "Snatch.", "Superbad")) %>%
  pandoc.table(emphasize.strong.cols = 3, missing="?")

@@

```{r results='asis'}
genders %>%
  sample_n(10) %>%
  select(first_name, dob_year, gender) %>%
  pandoc.table(emphasize.strong.cols = 3)
```

So generally this is looking pretty, pretty good. Things get dicey with unisex names like Sandy; however, the package must know that most Sandys born in 1937 were female so we'll have to trust it here. Finally we'll need to join this information back to the individual data set.

@@

gender_prop_finding = create_finding("The average age of a female actor at the time of film production is 28!")

@@

So henceforth we'll treat directors as a single bloc and you can assume that they're male statistically speaking.

# I starred in or directed a film; how old am I likely to be?

Well, as it turns out, it depends. We'll group individuals in groups: directors (predominantly male as seen above), male actors, and female actors. For all these individuals we calculate their age at the time a film they were in was realeased and plot the densities within each group.

Switch between the denisty (proportion appearing at one speicific age) and cumulative distribution (proportion appearing at **or before** at specific age) views to better understand the data.

@@

@@

> Insert infographic thing

@@

To investigate this hypothesis we must make a couple of assumptions. Since we don't have individual actor / director performance metrics (like we have in sports) we'll have to assume an actor or director can have a material affect on the quality of the movie they appear in; there are obviously many other factors but this seems like a relatively safe assumption.

@@

<iframe width="560" height="315" src="https://www.youtube.com/embed/qIKPJlKHKxg" frameborder="0" allowfullscreen></iframe>

@@

> INSERT EXAMPLES FROM A FEW ROTTEN TOMATOES PAGES

## Methodology

There's a few other things we want to control for here. I don't want to plot raw film critical rating because we might be prone to idiosyncracies of groups or ages - high performing directors might generally wait until they feel experienced enough to direct their first film as an example.

Instead we'll plot, for each individual and film, how well they have perfomed relative to that individual's mean lifetime score (the average tomatometer of all that individual's films). This gives us a common scale on which to compare.

To illustrate this approach let's look at Denzel Washington's films. Denzel has a healthy mean tomatometer of 68%.

```{r echo=FALSE, results='asis'}
master %>% 
  filter(
    !(role %in% c("actor_4","actor_5", "actor_6")),
    name == "Denzel Washington",
    between(age_at_production, 15,85)
  ) %>%
  group_by(name) %>% 
  mutate(
    average_tomato  = round(mean(tomatometer))
  ) %>% 
  ungroup() %>%
  mutate(
    tomato_index = (tomatometer - average_tomato) / average_tomato
  ) %>%
  sample_n(10) %>%
  arrange(age_at_production) %>%
  select(title, name, year, age_at_production, tomatometer, average_tomato, tomato_index) %>%
  mutate(tomato_index = scales::percent(tomato_index)) %>%
  pandoc.table(emphasize.strong.cols = 7)

@@

So based on this 10 film sample Denzel had a solid 40s where his films (of the likes **Hurricane** and **Courage Under Fire**) average a rotten tomatoes rating about 20% above his career average. He looks like he dropped off after 50 with a few flops like **Déjà Vu** (which I didn't mind) and **2 Guns**. Look I'm not calling him washed but...

## Results

So we use this approach and plot out the median tomato index by age at production, giving us an idea of the *average performance above or below a career average* at each age.

@@

# The prime suspects

actors = c("Robert De Niro", "Harvey Keitel", "Christopher Walken", "Bruce Willis", "Susan Sarandon", "Jeff Bridges", "Robin Williams", "Julianne Moore", "Steve Buscemi", "John Malkovich", "Meryl Streep", "Robert Redford")

}) %>% hw_grid(ncol = 3, rowheight = 400) %>% browsable()

@@

>>>>

.finding {
  background-color:#000;
  height: 125px;
  color: #FFF;
  display: flex;
  border-radius: 25px;
  align-items: center;
  text-align:center;
}

.finding .sidebar { 
  padding-top:15px;
  width: 30%;
  font-weight: bold;
  font-size: 18px;
}

.finding .body {
  width: 70%;
  text-align:left;
  font-size: 16px;
  padding-right:10px
}

.finding .body p {
  margin-bottom:0;
}

#the-half-list-of-acting-talent img {
  height: 200px; 
  display:block; 
  margin: 0 auto;
}

@@

  html = 
glue::glue("<br><br><div class='finding'>

<div class='sidebar'>
  <img src='https://github.com/TMBish/lab_posts/blob/master/00_utilities/lightbulb.png?raw=true'>
  <span> Finding!</span>
</div>

<div class='body'> 
  <span> {finding_text} </span> 
</div>

</div><br><br>")

@@

dict %>% pandoc.table(style = "rmarkdown")

@@

films %>% 
  filter(title %in% c("The Departed", "Titanic", "No Country for Old Men", "Snatch.", "Superbad")) %>%
  pandoc.table(emphasize.strong.cols = 3, missing="?", style = "rmarkdown", justify="left",split.cells=15)

@@

```{r results='asis'}
genders %>%
  sample_n(10) %>%
  select(first_name, dob_year, gender) %>%
  pandoc.table(emphasize.strong.cols = 3, style = "rmarkdown", justify="left", split.cells=15)
```

So generally this is looking pretty, pretty good. Things get dicey with unisex names like Sandy; however, the package must know that most Sandys born in 1937 were female so we'll have to trust it here.

@@

gender_prop_finding = create_finding("Approximately 80% of all directors are male, and 85% of popular films are directed by a man.")

@@

So we'll treat directors as a single group that, statistically speaking, are very much likely to be males.

# What's a typical career length?

Well, as it turns out, it depends on which one of the above groups you're in. To remind you, our groups are: directors (predominantly male as seen above), male actors, and female actors. For all these individuals we calculate their age at the time they were involved in a realeased film and plot the distribution within each group.

Click on the male and director series to compare with the female one shown below. Additionally, switch between the denisty view (proportion appearing at one speicific age) and cumulative distribution view (proportion appearing at **or before** at specific age) to better understand the data.

@@

age_atprod_finding = create_finding("64% of female roles are given to actors under the age of 40 compared with 42% of male roles. Only 28% of popular films are directed by a someone under 40.")

@@

@@

Our ultimate question is: do actors and directors improve or deteriorate as they get older. To investigate this hypothesis we must make a couple of assumptions. Since we don't have individual actor or director performance metrics (like we have in sports) we'll have to assume an actor or director can have a material affect on the quality of the moviesthey appear in. There are obviously many other factors that go into the quality of a film but this seems like a relatively safe assumption.

@@

<iframe width="560" height="315" style="display:block;margin: 0 auto;" src="https://www.youtube.com/embed/qIKPJlKHKxg" frameborder="0" allowfullscreen></iframe>

@@

### Snatch

![](snatch.png)

### No Country For Old Men

![](no-country-for-old-men.png)

### The Departed

![](the-departed.png)

So we'll limit this dataset to the top 3 actors in each film.

There's a few other things we want to control for here. I don't necessarily want to plot raw film critical rating because this might be prone to idiosyncracies within groups or ages - for example, high performing directors might generally wait until they feel experienced enough to direct their first film.

Instead we'll plot, for each individual and film, how well they have perfomed relative to that individual's career average (the average tomatometer of all that individual's films). To illustrate this approach let's look at Denzel Washington's films. Denzel has a healthy mean tomatometer of 68%.

```{r echo=FALSE, results='asis'}

master %>% 
  filter(
    !(role %in% c("actor_4","actor_5", "actor_6")),
    name == "Denzel Washington",
    between(age_at_production, 15,85)
  ) %>%
  group_by(name) %>% 
  mutate(
    average_tomato  = round(mean(tomatometer))
  ) %>% 
  ungroup() %>%
  mutate(
    tomato_index = (tomatometer - average_tomato) / average_tomato
  ) %>%
  sample_n(10) %>%
  arrange(age_at_production) %>%
  select(title, year, age_at_production, tomatometer, average_tomato, tomato_index) %>%
  mutate(tomato_index = scales::percent(tomato_index)) %>%
  pandoc.table(emphasize.strong.cols = 6, style = "rmarkdown", justify="left")

@@

So based on this 10 film sample Denzel had a solid early 40s where his films (of the likes **Hurricane** and **Courage Under Fire**) average a rotten tomatoes rating about 20% above his career average. He looks like he dropped off after 50 with a few flops like **Déjà Vu** (which I didn't mind) and **2 Guns**. Look I'm not calling him washed but he hasn't done anything good for a while.

## Prime Suspects

To help visualise this approach let's take 12 of the top 20 actors by volume and visualise average their film rating performance by age. De Niro is a great example of someone who starred in increasingly poor films as he got older - clearly visible in the chart below.. Obviously this probably has more to do with role selection than deteriorating skills (we'll discuss this later) but it's interesting to visualise nonetheless.

actors = c("Robert De Niro", "Harvey Keitel", "Christopher Walken", "Bruce Willis", "Susan Sarandon", "Jeff Bridges", "Robin Williams", "Julianne Moore", "Denzel Washington", "John Malkovich", "Meryl Streep", "Robert Redford")

}) %>% hw_grid(ncol = 3, rowheight = 350) %>% browsable()

## The holistic trend

So let's extend this approach and plot out the median tomato index by age at production, giving us an idea of the *average performance above or below a career average* at each age.

@@

actor_prime_finding = create_finding("Actors and Directors work in their most critically acclaimed films - about 10% above their career average - between 27 and 35 years old.")

actor_prime_finding

@@

# Caveats & Conclusions

While blah blah blah
